Cluster efficient pangenome graph construction with nf-core/pangenome

### 2024-05-13 manuscript/main.tex
feedback SN

	\abstract{
	\textbf{Motivation:}
	Pangenome graphs offer a comprehensive way of capturing genomic variability across multiple genomes. 
	However, current construction methods often introduce biases, excluding complex sequences or relying on references. 
	The PanGenome Graph Builder (PGGB) addresses these issues. Yet, its bash implementation hinders deployment, resource optimization, and scalability for large-scale graph construction. 
	\\
	\textbf{Results:}
	To overcome these limitations, we present \textit{nf-core/pangenome}, a reference-unbiased approach implemented in Nextflow following nf-core's best practices. 
	Leveraging biocontainers ensures portability and seamless deployment in HPC environments. 
	Unlike PGGB, nf-core/pangenome distributes alignments across cluster nodes, enabling scalability. 
	Demonstrating its efficiency, we constructed pangenome graphs for 1000 human chromosome 19 haplotypes and 2146 \textit{E. coli} sequences, achieving a two to threefold speedup compared to PGGB without increasing greenhouse gas emissions. \\
	\textbf{Availability:}
	nf-core/pangenome is released under the MIT open-source license, available on \href{https://github.com/nf-core/pangenome}{GitHub} and \href{https://zenodo.org/doi/10.5281/zenodo.8202636}{Zenodo}, with documentation accessible at \href{https://nf-co.re/pangenome/1.1.2/docs/usage}{https://nf-co.re/pangenome/1.1.2/docs/usage}. \\
	\textbf{Contact:} \href{simon.heumos@qbic.uni-tuebingen.de}{simon.heumos@qbic.uni-tuebingen.de}, \href{sven.nahnsen@qbic.uni-tuebingen.de}{sven.nahnsen@qbic.uni-tuebingen.de} \\
%	Pangenome graphs can encode the entire genomic variability between multiple genomes. Current pangenome graph construction methods exclude complex sequences or are reference-based, leading to reference, order, or orientation bias. This was addressed by the PanGenome Graph Builder (PGGB) pipeline. 
%	However, PGGB’s bash implementation limits its ease of deployment, optimal use of compute resources, and cluster scalability, making it impractical to build very large pangenome graphs.
%	\\
%	\textbf{Results:}
%	We implemented \textit{nf-core/pangenome}, a reference-unbiased approach to construct pangenome graphs. 
%	Mirroring PGGB, it iteratively refines an all-to-all whole-genome alignment graph that allows to explore sequence conservation and variation, infer phylogeny, and identify recombination events. 
%	nf-core/pangenome is implemented in Nextflow and follows the nf-core best practice development guidelines. 
%	Providing all software dependencies in biocontainers makes the pipeline portable and easy to install on high-performance computing environments. 
%	In contrast to PGGB, this allows nf-core/pangenome to distribute the quadratically complex all-to-all base-level alignments across nodes of a cluster. 
%	%Evaluating 1024 \textit{E. coli} haplotypes, the time spent on base-pair level alignments is reduced linearly with an increase in alignment problem chunks. 
%	To demonstrate the scalability of nf-core/pangenome, we built pangenome graphs of 1000 chromosome 19 human haplotypes, and of 2146 \textit{E. coli} sequences. nf-core/pangenome was two to three times faster compared to PGGB while not increasing the greenhouse gas emissions.
%	\\
%	\textbf{Availability:}
%	nf-core/pangenome is published as free software under the MIT open-source license. Source code can be downloaded from \url{https://github.com/nf-core/pangenome} and the documentation is accessible at \url{https://nf-co.re/pangenome/1.1.2/docs/usage}. 
%	Each release is archived on Zenodo \texttt{\href{https://zenodo.org/doi/10.5281/zenodo.8202636}{10.5281/zenodo.8202636}}. \\
%	\textbf{Contact:} \href{simon.heumos@qbic.uni-tuebingen.de}{simon.heumos@qbic.uni-tuebingen.de}, \href{sven.nahnsen@qbic.uni-tuebingen.de}{sven.nahnsen@qbic.uni-tuebingen.de} \\
	%\textbf{Supplementary information:} Supplementary data are available online. %at \textit{Bioinformatics} online.
	}

@@

	Current pangenome graph construction methods exclude complex sequences or are reference-biased \citep{Chin2023, Minkin2016}.
	%, are tree-guided, or reference-based \citep{Li:2020, Hickey2023} leading to reference, order, or orientation bias. 
	%Although whole genome scaling approaches for unbiased pangenome graph construction \citep{Chin2023, Minkin2016} exist, their reliance on k-mer-based data structures often leads to unwanted complexity for downstream analysis. 
	One recent approach that overcomes such limitations is the PanGenome Graph Builder (PGGB) pipeline \citep{Garrison2023}. 
	PGGB iteratively refines an all-to-all whole-genome alignment graph that lets us explore sequence conservation and variation, infer phylogeny, and identify recombination events. 
	PGGB was already extensively evaluated \citep{Garrison2023, Andreace2023} and applied to build the first draft human pangenome reference \citep{Liao2023}.
	However, PGGB is implemented in bash: This (a) makes it difficult to deploy on HPC systems, (b) does not allow for a fine granular tuning of computing resources for different steps of the pipeline \citep{Sztuka2024}, and (c) limits its cluster scalability because PGGB can only use the resources of one node.

>>>>

	\abstract{
	\textbf{Motivation:}
	Pangenome graphs offer a comprehensive way of capturing genomic variability across multiple genomes. 
	However, current construction methods often introduce biases, excluding complex sequences or relying on references. 
	The PanGenome Graph Builder (PGGB) addresses these issues. 
	To date, though, there is no state-of-the-art pipeline allowing for easy deployment, efficient and dynamic use of available resources, and scalable usage at the same time.
	%Yet, its bash implementation hinders deployment, resource optimization, and scalability for large-scale graph construction. 
	\\
	\textbf{Results:}
	To overcome these limitations, we present \textit{nf-core/pangenome}, a reference-unbiased approach implemented in Nextflow following nf-core's best practices. 
	Leveraging biocontainers ensures portability and seamless deployment in HPC environments. 
	%SN: in any compute infrastructure, such as HPCs or clouds
	Unlike PGGB, nf-core/pangenome distributes alignments across cluster nodes, enabling scalability. 
	Demonstrating its efficiency, we constructed pangenome graphs for 1000 human chromosome 19 haplotypes and 2146 \textit{E. coli} sequences, achieving a two to threefold speedup compared to PGGB without increasing greenhouse gas emissions. \\
	\textbf{Availability:}
	nf-core/pangenome is released under the MIT open-source license, available on \href{https://github.com/nf-core/pangenome}{GitHub} and \href{https://zenodo.org/doi/10.5281/zenodo.8202636}{Zenodo}, with documentation accessible at \href{https://nf-co.re/pangenome/1.1.2/docs/usage}{https://nf-co.re/pangenome/1.1.2/docs/usage}. \\
	\textbf{Contact:} \href{simon.heumos@qbic.uni-tuebingen.de}{simon.heumos@qbic.uni-tuebingen.de}, \href{sven.nahnsen@qbic.uni-tuebingen.de}{sven.nahnsen@qbic.uni-tuebingen.de} \\
%	Pangenome graphs can encode the entire genomic variability between multiple genomes. Current pangenome graph construction methods exclude complex sequences or are reference-based, leading to reference, order, or orientation bias. This was addressed by the PanGenome Graph Builder (PGGB) pipeline. 
%	However, PGGB’s bash implementation limits its ease of deployment, optimal use of compute resources, and cluster scalability, making it impractical to build very large pangenome graphs.
%	\\
%	\textbf{Results:}
%	We implemented \textit{nf-core/pangenome}, a reference-unbiased approach to construct pangenome graphs. 
%	Mirroring PGGB, it iteratively refines an all-to-all whole-genome alignment graph that allows to explore sequence conservation and variation, infer phylogeny, and identify recombination events. 
%	nf-core/pangenome is implemented in Nextflow and follows the nf-core best practice development guidelines. 
%	Providing all software dependencies in biocontainers makes the pipeline portable and easy to install on high-performance computing environments. 
%	In contrast to PGGB, this allows nf-core/pangenome to distribute the quadratically complex all-to-all base-level alignments across nodes of a cluster. 
%	%Evaluating 1024 \textit{E. coli} haplotypes, the time spent on base-pair level alignments is reduced linearly with an increase in alignment problem chunks. 
%	To demonstrate the scalability of nf-core/pangenome, we built pangenome graphs of 1000 chromosome 19 human haplotypes, and of 2146 \textit{E. coli} sequences. nf-core/pangenome was two to three times faster compared to PGGB while not increasing the greenhouse gas emissions.
%	\\
%	\textbf{Availability:}
%	nf-core/pangenome is published as free software under the MIT open-source license. Source code can be downloaded from \url{https://github.com/nf-core/pangenome} and the documentation is accessible at \url{https://nf-co.re/pangenome/1.1.2/docs/usage}. 
%	Each release is archived on Zenodo \texttt{\href{https://zenodo.org/doi/10.5281/zenodo.8202636}{10.5281/zenodo.8202636}}. \\
%	\textbf{Contact:} \href{simon.heumos@qbic.uni-tuebingen.de}{simon.heumos@qbic.uni-tuebingen.de}, \href{sven.nahnsen@qbic.uni-tuebingen.de}{sven.nahnsen@qbic.uni-tuebingen.de} \\
	%\textbf{Supplementary information:} Supplementary data are available online. %at \textit{Bioinformatics} online.
	}

@@

	Current pangenome graph construction methods exclude complex sequences or are reference-biased \citep{Chin2023, Minkin2016}.
	%, are tree-guided, or reference-based \citep{Li:2020, Hickey2023} leading to reference, order, or orientation bias. 
	%Although whole genome scaling approaches for unbiased pangenome graph construction \citep{Chin2023, Minkin2016} exist, their reliance on k-mer-based data structures often leads to unwanted complexity for downstream analysis. 
	One recent approach that overcomes such limitations is the PanGenome Graph Builder (PGGB) pipeline \citep{Garrison2023}. 
	PGGB iteratively refines an all-to-all whole-genome alignment graph that lets us explore sequence conservation and variation, infer phylogeny, and identify recombination events. 
	PGGB was already extensively evaluated \citep{Garrison2023, Andreace2023} and applied to build the first draft human pangenome reference \citep{Liao2023}.
	However, PGGB is implemented in bash: This (a) makes it difficult to deploy on HPC systems, (b) does not allow for a fine granular tuning of computing resources for different steps of the pipeline \citep{Sztuka2024}, and (c) limits its cluster scalability because PGGB can only use the resources of one node.
	%SN: Given the high resource requirements of large-scale pangenomes, these limitations are significantly limiting broad applications.
